---
title: Figure 1. BugSigDB overview
vignette: >
  % \VignetteIndexEntry{Figure 1}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output:
  html_document:
    mathjax: null   
---

Setup:

```{r setup, message=FALSE, warning=FALSE}
library(bugsigdbr)
library(ggpubr)
library(lubridate)
library(BugSigDBStats)
library(ComplexHeatmap)
```

Panel A: curation output over time

```{r}
dat <- bugsigdbr::importBugSigDB(version = "10.5281/zenodo.5904281", cache = FALSE)
dat[,"Curated date"] <- as.character(lubridate::dmy(dat[,"Curated date"]))
dbm <- BugSigDBStats::plotProgressOverTime(dat)
df <- reshape2::melt(dbm)
df$Var1 <- ifelse(df$Var1 == "cdbm", "signatures", "papers")

p <- ggbarplot(df, x = "Var2", y = "value", fill = "Var1", 
               color = "Var1", palette = "Paired",
               position = position_dodge(0.9), orientation = "horiz",
               ggtheme = theme_bw(), legend = "top", xlab = "", ylab = "",
               legend.title = "")
```

Panel B and C: papers per country and body site

```{r}
ppc <- BugSigDBStats::paperStat(dat, "Location of subjects")
p1 <- BugSigDBPaper::plotStatsPerCountry(ppc, fixed = FALSE)
bbc <- BugSigDBStats::paperStat(dat, "Body site")
p2 <- BugSigDBPaper::plotStatsPerBodySite(bbc, sex = "female") + 
      theme(legend.position = "none")
```

Arrange panels A-C:

```{r}
p12 <- ggarrange(p, p1, p2, widths = c(1.5, 2.5, 1),
                 labels = c("A", "B", "C"), ncol = 3)
```

Panel D: most frequent taxa

```{r}
msc <- bugsigdbr::getSignatures(dat, tax.id.type = "metaphlan")
(nuniq <- length(unique(unlist(msc))))
msc.tab <- BugSigDBStats::getMostFrequentTaxa(dat, nuniq, tax.id.type = "metaphlan")
msc.tab[5] <- msc.tab[5] + 5
msc.tab[8] <- msc.tab[8] - 5
msc.tab[10] <- msc.tab[10] - 8
df.text <- data.frame(x = as.vector(msc.tab)[1:10], y = 1, label = names(msc.tab)[1:10])
ggp <- gghistogram(as.vector(msc.tab), bins = 20,
            xlab = "number of signatures",
            ylab = "number of microbes",
            fill = "#00AFBB", 
            ggtheme = theme_bw()) + 
            theme(axis.title = element_text(size = 10))
p4 <- ggp + geom_text(data = df.text,
                aes(x = x, y = y, label = label),
                angle = 90, hjust = 0, nudge_y = 5, size = 3)
```

Panel E: microbe set size

```{r}
p5 <- gghistogram(lengths(msc), bins = 30, ylab="number of signatures", 
    xlab="signature size", fill = "#00AFBB", 
    ggtheme = theme_bw()) +
    theme(axis.title = element_text(size = 10))
p45 <- ggarrange(p4, p5, nrow = 2, labels = c("D", "E"), vjust = 0.5)
```

Conditions on signature level - based on pre-defined categories:

```{r}
data.dir <- system.file("extdata", package = "BugSigDBPaper")
tab.file <- file.path(data.dir, "condition2category.txt")
tab <- read.csv(tab.file, header = FALSE)
dat[,"Condition"] <- tolower(dat[,"Condition"])
dat <- subset(dat, Condition %in% tab[,1])

cond2cat <- tab[,2]
names(cond2cat) <- tab[,1]

cats <- cond2cat[dat[,"Condition"]]
x <- table(unname(cats))
categs <- names(sort(x,decreasing = TRUE))
p6 <- BugSigDBStats::plotComposition(sort(x, decreasing = TRUE), "condition") + 
      theme(legend.title = element_text(size = 8), legend.text = element_text(size = 6))

leg <- ggpubr::get_legend(p6)
p6 <- BugSigDBStats::plotComposition(sort(x, decreasing = TRUE), "condition") +
      theme(legend.position = "none", text = element_text(size = 8))
```

Conditions on taxon level:

```{r}
fts <- names(msc.tab)[1:10]
ftss <- sub("^[a-z]__", "", fts) 
fts.stats <- lapply(ftss, BugSigDBStats::getTaxonStats,
                    dat = dat, column = "Condition")
for(i in seq_along(fts.stats)) names(fts.stats[[i]]) <- cond2cat[names(fts.stats[[i]])]
.f <- function(stat) vapply(unique(names(stat)), function(n) sum(stat[names(stat) == n]), integer(1))
fts.stats <- lapply(fts.stats, .f)
for(i in seq_along(fts.stats)) fts.stats[[i]] <- fts.stats[[i]][categs]
mat <- do.call(cbind, fts.stats)
colnames(mat) <- fts
rownames(mat)[10] <- "medical treatment"
mat[is.na(mat)] <- 0
mat2 <- sweep(mat, 2, colSums(mat), `/`)
```

```{r}
df <- reshape2::melt(mat)
colnames(df) <- c("condition", "taxon", "count")
p7 <- ggpubr::ggbarplot(df, x = "taxon", y = "count", fill = "condition",
               color = "condition", palette = "npg",
               orientation = "horiz",
               ggtheme = theme_bw(),# legend = "none",
               ylab = "number of signatures", xlab = "")
p67 <- ggpubr::ggarrange(p6, p7, legend.grob = leg, labels = c("F", "G"))
p47 <- ggpubr::ggarrange(p45, p67, widths = c(1,2))
```

Proportion tests of Bacteroides with certain condition categoires when compared
to their background distribution in the database:

```{r}
prop.test(c(mat["metabolic disease", "g__Bacteroides"], x["metabolic disease"]),
          c(sum(mat[,"g__Bacteroides"]),sum(x)))
prop.test(c(mat["antibiotics", "g__Bacteroides"], x["antibiotics"]),
          c(sum(mat[,"g__Bacteroides"]),sum(x)))
```

Putting it all together:

```{r, fig.width = 10, fig.height = 7}
ggpubr::ggarrange(p12, p47, nrow = 2, heights = c(1,1.3))
```

```{r, eval = FALSE}
png("Figure1.png", width = 3375, height = 2109, res = 300)
ggpubr::ggarrange(p12, p47, nrow = 2, heights = c(1,1.3))
dev.off()
```

Supplementary Figure: publication date of the curated papers:

Number of papers and signatures curated:

```{r nrPapers}
pmids <- unique(dat[,"PMID"])
length(pmids)
nrow(dat)
```

Publication date of the curated papers:

```{r pubDate}
pmids <- pmids[!is.na(pmids)]
pubyear1 <- BugSigDBStats::pmid2pubyear(pmids[1:361])
pubyear2 <- BugSigDBStats::pmid2pubyear(pmids[362:length(pmids)])
pubyear <- c(pubyear1, pubyear2)
head(cbind(pmids, pubyear))
```

```{r pubDate2}
tab <- table(pubyear)
tab <- tab[-length(tab)]
tab <- tab[order(as.integer(names(tab)))]
df <- data.frame(year = names(tab), papers = as.integer(tab))
ggpubr::ggbarplot(df, x = "year", y = "papers", 
                  label = TRUE, fill = "steelblue",
                  ggtheme = theme_bw())
```

Supplementary Figure: distribution of taxonomic levels:

```{r taxLevel}
tab <- BugSigDBStats::getTaxLevelDistribution(msc)
tab <- tab[-1]
df <- data.frame(level = names(tab), count = as.integer(tab))
ggbarplot(df, x = "level", y = "count",
          xlab = "Taxonomic level", ylab = "Number of unique taxa", 
          label = TRUE, fill = "steelblue", ggtheme = theme_bw())
```
