---
title: Figure 4. Microbe co-occurrence 
vignette: >
  % \VignetteIndexEntry{Figure 4}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output:
  html_document:
    mathjax: null   
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
library(bugsigdbr)
library(curatedMetagenomicData)
library(ggalluvial)
library(ggpubr)
library(BugSigDBStats)
library(ComplexHeatmap)
```

# Microbe co-occurrence in fecal samples from BugSigDB

```{r}
#bsdb <- bugsigdbr::importBugSigDB(version = "1bf4e1b", cache = FALSE)
bsdb <- bugsigdbr::importBugSigDB(version = "10.5281/zenodo.5904281", cache = FALSE)
bsdb.feces <- subset(bsdb, `Body site` == "feces")
```

```{r coocc, fig.width = 10, fig.height = 10}
cooc.mat <- BugSigDBStats::microbeHeatmap(bsdb.feces, tax.level = "genus")
antag.mat <- BugSigDBStats::microbeHeatmap(bsdb.feces,
                                           tax.level = "genus",
                                           antagonistic = TRUE)
```

Get the top 20 genera most frequently reported as differentially abundant:

```{r}
sigs.feces <- bugsigdbr::getSignatures(bsdb.feces, tax.id.type = "taxname",
                            tax.level = "genus", exact.tax.level = FALSE)
top20 <- sort(table(unlist(sigs.feces)), decreasing = TRUE)[1:20]
top20
```

Subset heatmaps to the top 20 genera most frequently reported as differentially abundant:

```{r}
all(names(top20) %in% rownames(cooc.mat))
cooc.mat <- cooc.mat[names(top20), names(top20)]
all(names(top20) %in% rownames(antag.mat))
antag.mat <- antag.mat[names(top20), names(top20)]
```

Distinguish by direction of abundance change (increased / decreased):

```{r}
# increased
sub.bsdb.feces <- subset(bsdb.feces, `Abundance in Group 1` == "increased")
sigs.feces.up <- getSignatures(sub.bsdb.feces, tax.id.type = "taxname",
                               tax.level = "genus", exact.tax.level = FALSE)
top20.up <- table(unlist(sigs.feces.up))[names(top20)]
top20.up

# decreased
sub.bsdb.feces <- subset(bsdb.feces, `Abundance in Group 1` == "decreased")
sigs.feces.down <- getSignatures(sub.bsdb.feces, tax.id.type = "taxname",
                                 tax.level = "genus", exact.tax.level = FALSE)
top20.down <- table(unlist(sigs.feces.down))[names(top20)]
top20.down
```

Plot the heatmap:

```{r mheat, fig.width = 10, fig.height = 10}
# annotation
mat <- matrix(nc = 2, cbind(top20.up, top20.down))
bp <- ComplexHeatmap::anno_barplot(mat, gp = gpar(fill = c("#D55E00", "#0072B2"),
                                                  col = c("#D55E00", "#0072B2")),
                                   height = unit(2, "cm"))
banno <- ComplexHeatmap::HeatmapAnnotation(`Abundance in Group 1` = bp)

lgd_list <- list(
    Legend(labels = c("increased", "decreased"),
           title = "Abundance in Group 1",
           type = "grid",
           legend_gp = gpar(col = c("#D55E00", "#0072B2"), fill = c("#D55E00", "#0072B2"))))

# same direction
# lcm <- sweep(cooc.mat, 2, matrixStats::colMaxs(cooc.mat), FUN = "/")
# we need to dampen the maximum here a bit down,
# otherwise 100% self co-occurrence takes up a large fraction of the colorscale,
sec <- apply(cooc.mat, 2, function(x) sort(x, decreasing = TRUE)[2])
cooc.mat2 <- cooc.mat
for(i in 1:ncol(cooc.mat2)) cooc.mat2[i,i] <- min(cooc.mat2[i,i], 1.4 * sec[i])
lcm <- sweep(cooc.mat2, 2, matrixStats::colMaxs(cooc.mat2), FUN = "/")
col <- circlize::colorRamp2(c(0,1), c("#EEEEEE", "red"))
ht1 <- ComplexHeatmap::Heatmap(lcm,
                               col = col,
                               name = "Relative frequency top",
                               cluster_columns = FALSE,
                               row_km = 3,
                               row_title = "same direction",
                               column_names_rot = 45,
                               row_names_gp = gpar(fontsize = 8),
                               column_names_gp = gpar(fontsize = 8))

# opposite direction
acm <- sweep(antag.mat, 2, matrixStats::colMaxs(antag.mat), FUN = "/")
col <- circlize::colorRamp2(c(0,1), c("#EEEEEE", "blue"))
ht2 <- ComplexHeatmap::Heatmap(acm,
                               col = col,
                               name = "Relative frequency bottom",
                               cluster_columns = FALSE,
                               row_title = "opposite direction",
                               row_km = 3,
                               column_names_rot = 45,
                               row_names_gp = gpar(fontsize = 8),
                               column_names_gp = gpar(fontsize = 8))

# phylum
sfp <- bugsigdbr::getSignatures(bsdb.feces, tax.id.type = "metaphlan",
                                tax.level = "genus", exact.tax.level = FALSE)
sfp20 <- sort(table(unlist(sfp)), decreasing = TRUE)[1:20]
uanno <- bugsigdbr::extractTaxLevel(names(sfp20),
                                    tax.id.type = "taxname",
                                    tax.level = "phylum",
                                    exact.tax.level = FALSE)
phyla.grid <- seq_along(unique(uanno))
panno <- ComplexHeatmap::HeatmapAnnotation(phylum = uanno)

uanno <- matrix(uanno, nrow = 1)
colnames(uanno) <- names(top20)
pcols <- c("#CC79A7", "#F0E442", "#009E73", "#56B4E9", "#E69F00")
uanno <- ComplexHeatmap::Heatmap(uanno, name = "Phylum",
                                 col = pcols[phyla.grid],
                                 cluster_columns = FALSE,
                                 column_names_rot = 45,
                                 column_names_gp = gpar(fontsize = 8))

# put everything together
ht_list <- ht1 %v% banno %v% ht2 %v% uanno
ComplexHeatmap::draw(ht_list, annotation_legend_list = lgd_list, merge_legend = TRUE)

decorate_annotation("Abundance in Group 1", {
    grid.text("# signatures", x = unit(-1, "cm"), rot = 90, just = "bottom", gp = gpar(fontsize = 8))
    grid.text("*", x = unit(2.45, "cm"), y = unit(1.2, "cm"))
    grid.text("*", x = unit(5.18, "cm"), y = unit(1, "cm"))
    grid.text("*", x = unit(6.55, "cm"), y = unit(0.95, "cm"))
    grid.text("*", x = unit(8.6, "cm"), y = unit(0.85, "cm"))
    grid.text("*", x = unit(10, "cm"), y = unit(0.7, "cm"))
    grid.text("*", x = unit(10.7, "cm"), y = unit(0.7, "cm"))
})
```

# Contrasting with microbe prevalence in healthy samples

Get healthy samples from curatedMetagenomicData:

```{r}
healthy <- subset(curatedMetagenomicData::sampleMetadata, disease == "healthy")
ind <- apply(healthy, 2, function(x) !all(is.na(x)))
healthy <- healthy[,ind]
```

Stratify by age:

```{r}
adult <- subset(healthy, age_category %in% c("adult", "senior"))
```

Helper function for obtaining signature passing prevalence threshold:

```{r}
prevalencecutoff <- 0.000001
.returnSig <- function(obj, threshold = 0, prevalence = 0.000001)
{
    fractionpassing.logical <- rowSums(obj > threshold) / ncol(obj)
    rows.passing <- fractionpassing.logical > prevalence
    return(fractionpassing.logical[rows.passing])
}
```

Obtain stool samples:

```{r, message = FALSE}
cmd.stool <- subset(adult, body_site == "stool")
cmd.stool <- curatedMetagenomicData::returnSamples(cmd.stool,
                                                   "relative_abundance",
                                                   rownames = "short")
```

Pull out prevalence on genus level:

```{r}
cmd.stool.byranks <- mia::splitByRanks(cmd.stool)
cmd.stool.genus <- assay(cmd.stool.byranks[["genus"]])
prev.stool.genus <- .returnSig(cmd.stool.genus, prevalence = prevalencecutoff)
```

Restrict to top 20 genera displayed in Heatmap: 

```{r}
prev.top20 <- prev.stool.genus[names(top20)]
prev.top20
```

Add prevalence as barplot:

```{r, fig.width = 10, fig.height = 10}
prev <- ComplexHeatmap::anno_barplot(prev.top20,
                                     height = unit(2, "cm"), 
                                     gp = gpar(fill = "lightgrey"))
prev.anno <- ComplexHeatmap::HeatmapAnnotation(Prevalence = prev)
ht_list <- banno %v% prev.anno %v% ht1 %v% ht2 %v% uanno
cm <- grid.grabExpr({
    ComplexHeatmap::draw(ht_list, 
                         annotation_legend_list = lgd_list,
                         merge_legend = TRUE)

    decorate_annotation("Abundance in Group 1", {
        grid.text("# signatures", x = unit(-1, "cm"), 
                  rot = 90, just = "bottom", gp = gpar(fontsize = 8))
        grid.text("*", x = unit(1.7, "cm"), y = unit(1.2, "cm"))
        grid.text("*", x = unit(4.1, "cm"), y = unit(1, "cm"))
        grid.text("*", x = unit(4.6, "cm"), y = unit(0.95, "cm"))
        grid.text("*", x = unit(6.05, "cm"), y = unit(0.85, "cm"))
        grid.text("*", x = unit(7, "cm"), y = unit(0.7, "cm"))
        grid.text("*", x = unit(7.5, "cm"), y = unit(0.7, "cm"))
        grid.text("*", x = unit(9.45, "cm"), y = unit(0.6, "cm"))
    })
})
```

There seems to be a trend that high-prevalence genera tend to be reported with
decreased abundance and low-prevalence genera with increased abundance in the
diseased group:

```{r}
prop.up <- mat[,1] / rowSums(mat)
names(prop.up) <- names(top20)
cor.test(prev.top20, prop.up)
```

```{r}
res <- lm(prop.up ~ prev.top20)
summary(res)
#plot(res)
```

```{r}
df <- data.frame(prev = prev.top20, prop = prop.up)
prev.prop <- ggpubr::ggscatter(df, x = "prev", y = "prop", color = "firebrick",
          xlab = "Prevalence",
          ylab = "Proportion UP",
          add = "reg.line",
          conf.int = TRUE,     
          ggtheme = ggplot2::theme_bw(),
          legend = "none") +
          stat_cor(label.x.npc = "center")
prev.prop
```

# Contrasting with correlation in healthy samples

```{r}
cmd.stool.genus.top20 <- cmd.stool.genus[names(top20),]
cmat <- cor(t(cmd.stool.genus.top20), method = "spearman")
cmat[1:5,1:5]
```

Plot correlation heatmap:

```{r}
panno <- bugsigdbr::extractTaxLevel(names(sfp20),
                                    tax.id.type = "taxname",
                                    tax.level = "phylum",
                                    exact.tax.level = FALSE)
canno <- bugsigdbr::extractTaxLevel(names(sfp20), tax.id.type = "taxname", 
                                    tax.level = "class", exact.tax.level = FALSE)
ccols <- RColorBrewer::brewer.pal(6, "Set2")
names(ccols) <- unique(canno)
pcols <- pcols[1:4]
names(pcols) <- unique(panno)
# phylum = panno
# phylum = pcols
tanno <- ComplexHeatmap::HeatmapAnnotation(df = data.frame(class = canno),
                                           col = list(class = ccols))#,
                                           # annotation_legend_param = list(class = list(direction = "horizontal")))
diag(cmat) <- NA
corm <- ComplexHeatmap::Heatmap(cmat, top_annotation = tanno, name = "cor",
                                na_col = "lightgrey",
                                column_names_rot = 45,
                                row_names_gp = gpar(fontsize = 8),
                                column_names_gp = gpar(fontsize = 8))#,
                                #heatmap_legend_param = list(direction = "horizontal"))
corm <- grid.grabExpr(draw(corm, 
                           heatmap_legend_side = "bottom", 
                           annotation_legend_side = "bottom",
                           merge_legend = TRUE))
corm
diag(cmat) <- 1
```

Adding correlation to the co-occurence matrices:

```{r}
cell_fun <- function(j, i, x, y, width, height, fill)
{
        if(cmat[i,j] > 0 && cmat[i,j] < 1)
            grid.points(x, y, pch = 20, size = unit(cmat[i,j], "char"))
        if(cmat[i,j] == 1)
        {
            grid.rect(x = x, y = y, width = width, height = height, 
                      gp = gpar(col = NA, fill = "grey"))
            grid.points(x, y, pch = "*", gp = gpar(col = "white")) 
            
        }
            
}
```

```{r}
rcol <- circlize::colorRamp2(c(0,1), c("#EEEEEE", "red"))
ht1 <- ComplexHeatmap::Heatmap(lcm,
                               col = rcol,
                               cell_fun = cell_fun,
                               name = "Relative frequency top",
                               cluster_columns = FALSE,
                               row_km = 3,
                               row_title = "same direction",
                               column_names_rot = 45,
                               row_names_gp = gpar(fontsize = 8),
                               column_names_gp = gpar(fontsize = 8))
```

```{r}
cell_fun <- function(j, i, x, y, width, height, fill)
{
        if(cmat[i,j] < 0)
            grid.points(x, y, pch = 20, size = unit(cmat[i,j] * -1, "char"))
}
```

```{r}
bcol <- circlize::colorRamp2(c(0,1), c("#EEEEEE", "blue"))
ht2 <- ComplexHeatmap::Heatmap(acm,
                               col = bcol,
                               cell_fun = cell_fun,
                               name = "Relative frequency bottom",
                               cluster_columns = FALSE,
                               row_title = "opposite direction",
                               row_km = 3,
                               column_names_rot = 45,
                               row_names_gp = gpar(fontsize = 8),
                               column_names_gp = gpar(fontsize = 8))
```

```{r}
sgrid <- seq(-1, 1, by = 0.25)[-5]
bg <- rep(c(bcol(0.4), rcol(0.4)), each = 4)
ht_list <- banno %v% prev.anno %v% ht1 %v% ht2 %v% uanno
cor.leg <- ComplexHeatmap::Legend(labels = sgrid, 
                                  title = "Spearman", 
                                  type = "points", 
                                  pch = 20,
                                  size = unit(abs(sgrid), "char"),
                                  background = bg)
lgd_list <- c(lgd_list, cor.leg)

cm <- grid.grabExpr({
    ComplexHeatmap::draw(ht_list, 
                         annotation_legend_list = lgd_list,
                         merge_legend = TRUE)

    decorate_annotation("Abundance in Group 1", {
        grid.text("# signatures", x = unit(-1, "cm"), 
                  rot = 90, just = "bottom", gp = gpar(fontsize = 8))
        grid.text("*", x = unit(1.7, "cm"), y = unit(1.2, "cm"))
        grid.text("*", x = unit(4.1, "cm"), y = unit(1, "cm"))
        grid.text("*", x = unit(4.6, "cm"), y = unit(0.95, "cm"))
        grid.text("*", x = unit(6.05, "cm"), y = unit(0.85, "cm"))
        grid.text("*", x = unit(7, "cm"), y = unit(0.7, "cm"))
        grid.text("*", x = unit(7.5, "cm"), y = unit(0.7, "cm"))
        grid.text("*", x = unit(9.45, "cm"), y = unit(0.6, "cm"))
    })
})
```

Putting it all together:

```{r, fig.width = 11, fig.height = 11}
ri <- ggarrange(corm, prev.prop, heights = c(1.5, 1), nrow = 2, labels = c("B", "C"))
ggarrange(cm, ri, ncol = 2, widths = c(1.5,1), labels = c("A", ""))
```

# Identify conditions associated with certain genera  

```{r}
genus.bsdb.feces <- bugsigdbr::restrictTaxLevel(bsdb.feces, 
                                                tax.level = "genus",
                                                exact.tax.level = FALSE)
genus.bsdb.feces[["MetaPhlAn taxon names"]] <- lapply(genus.bsdb.feces[["MetaPhlAn taxon names"]], 
                                                      extractTaxLevel,
                                                      tax.id.type = "taxname")
```

A function to obtain stats on an individual taxon for a column of interest:

```{r}
getTaxonStats <- function(dat, taxon, 
                          column = "Condition",
                          direction = "UP")
{
    ind <- !is.na(dat[["Abundance in Group 1"]])
    dat <- dat[ind,]
    dir <- ifelse(direction == "UP", "increased", "decreased")
    dat <- subset(dat, `Abundance in Group 1` == dir) 
    ind <- vapply(dat[["MetaPhlAn taxon names"]], 
                  function(x) taxon %in% x, 
                  logical(1))
    tab <- table(dat[ind,column])
    tab <- sort(tab, decreasing = TRUE)
    return(tab)
}
```

```{r}
getTaxonStats(genus.bsdb.feces, "Enterococcus") 
```

A function to test association of a taxon with a category of a column of interest: 

```{r}
testAssociation <- function(dat, taxon, column = "Condition", 
                            category = "", 
                            direction = "UP")
{
    stats <- getTaxonStats(dat, taxon, column, direction)
    stopifnot(all(category %in% names(stats)))

    fg <- sum(stats[category])
    fg.total <- sum(stats)    

    dir <- ifelse(direction == "UP", "increased", "decreased")
    dat <- subset(dat, `Abundance in Group 1` == dir)

    bg <- sum(dat[,column] %in% category)
    bg.total <- nrow(dat)

    prop.test(c(fg, bg), c(fg.total, bg.total))
}
```

```{r}
testAssociation(genus.bsdb.feces, "Enterococcus", category = "schizophrenia")
```

Now go through all condition-taxon pairs for which there are at least five signatures
and test for over-representation, while distinguishing between up and down:

```{r}
tstats.up <- lapply(names(top20), getTaxonStats, dat = genus.bsdb.feces, direction = "UP")
tstats.down <- lapply(names(top20), getTaxonStats, dat = genus.bsdb.feces, direction = "DOWN")
names(tstats.up) <- names(tstats.down) <- names(top20)
tstats.up <- reshape2::melt(tstats.up)
tstats.down <- reshape2::melt(tstats.down)
tstats <- rbind(tstats.up, tstats.down)
tstats <- cbind(tstats, rep(c("UP", "DOWN"), c(nrow(tstats.up), nrow(tstats.down))))
colnames(tstats) <- c("Condition", "value", "Taxon", "Direction")
tstats <- subset(tstats, value > 4)
head(tstats)
```

Now test association using a proportion test:

```{r, warning = FALSE}
ps <- apply(tstats, 1, function(x) testAssociation(genus.bsdb.feces, 
                                                   taxon = x[3],
                                                   category = x[1],
                                                   direction = x[4])$p.value)
adjp <- p.adjust(ps, method = "BH")
tstats <- cbind(tstats, pvalue = ps, adj.pvalue = adjp)
tstats <- tstats[order(tstats$pvalue),]
head(tstats)
```

Associations with FDR < 0.05:

```{r}
sig.tstats <- subset(tstats, adj.pvalue < 0.05)
sig.tstats
```

Some cosmetics for plotting:

```{r}
sig.tstats$Direction <- factor(sig.tstats$Direction, levels = c("UP", "DOWN"))
sig.tstats$Condition <- as.character(sig.tstats$Condition)
ind <- sig.tstats$Condition != "COVID-19"
sig.tstats[ind,"Condition"] <- vapply(sig.tstats[ind,"Condition"], 
                                    stringr::str_to_sentence,
                                    character(1))
aggr <- aggregate(value ~ Condition, sig.tstats[,1:2], FUN = sum)
aggr <- aggr[order(aggr$value, decreasing = TRUE),]
sig.tstats$Condition <- factor(sig.tstats$Condition, levels = aggr$Condition)
aggr <- aggregate(value ~ Taxon, sig.tstats[,2:3], FUN = sum)
aggr <- aggr[order(aggr$value, decreasing = TRUE),]
sig.tstats$Taxon <- factor(sig.tstats$Taxon, levels = aggr$Taxon)
```


```{r}
gga <- ggplot(sig.tstats, aes(y = value, axis1 = Taxon, axis2 = Condition)) + 
    geom_alluvium(aes(fill = Direction)) + 
    scale_x_discrete(limits = c("Taxon", "Condition"), expand = c(.1, .1)) + 
    geom_stratum(width = 1/12, fill = "lightgrey", color = "darkgrey") + 
    geom_text(stat = "stratum", aes(label = after_stat(stratum))) + 
    ylab("Number of signatures") + 
    theme_bw() # + theme(legend.position = "none")
gga
```

