---
title: Figure 2. Signature similarity
vignette: >
  % \VignetteIndexEntry{Signature similarity}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output:
  html_document:
    mathjax: null   
---

Setup:

```{r, message = FALSE}
library(aplot)
library(bugsigdbr)
library(BugSigDBStats)
library(ComplexHeatmap)
library(ggpubr)
library(ggplotify)
library(ggtree)
library(ontologySimilarity)
```

# Helper functions

Function to pull out author and year of publication from a signature name:

```{r}
.getAuthorYear <- function(dat, sname)
{
    stopifnot(length(sname) == 1 && is.character(sname))
    id <- unlist(strsplit(sname, "_"))[1]
    stopifnot(grepl("^bsdb:", id))
    id <- sub("^bsdb:", "", id)
    id <- unlist(strsplit(id, "/"))
    sdat <- subset(dat, Study == paste0("Study ", id[1]) &
                        Experiment == paste0("Experiment ", id[2]) &
                        `Signature page name` == paste0("Signature ", id[3]))
    author <- sdat$Authors
    aspl <- unlist(strsplit(author, " "))[1]
    year <- sdat$Year
    cond <- sdat$Condition
    ay <- paste0(aspl, year)
    ay <- paste(ay, cond, sep = "_")
    return(ay)
}
```

Function to build a short name for a signature name:

```{r}
.getShortName <- function(sname)
{
    stopifnot(grepl("^bsdb:", sname))
    spl <- unlist(strsplit(sname, ":"))
    paste(spl[1], spl[2], sep = ":")
}
```

Function to parse out condition from signature names:

```{r}
.getCondition <- function(sigs)
{
    spl <- strsplit(sigs, "_")
    spl <- vapply(spl, `[`, character(1), x = 2)
    spl <- strsplit(spl, ":")
    spl <- vapply(spl, `[`, character(1), x = 1)
    return(spl)
}
```

Function to parse out direction from signature names:

```{r}
.getDirection <- function(sigs)
{
    spl <- strsplit(sigs, "_")
    spl <- vapply(spl, function(x) x[length(x)], character(1))
    return(spl)
}
```

# Obtain the data

```{r}
dat <- bugsigdbr::importBugSigDB(version = "10.5281/zenodo.5904281", cache = FALSE)
onto <- BugSigDBStats::getNcbiTaxonomyObo()
# onto <- ontologyIndex::get_ontology("http://purl.obolibrary.org/obo/ncbitaxon.obo")
```

# Semantic similarity on mixed taxonomic levels

First we look at the overall signature similarity matrix with a PCA/PCoA:

```{r}
ind <- lengths(dat[["NCBI Taxonomy IDs"]]) > 4
sub.dat <- dat[ind,]
sigs <- bugsigdbr::getSignatures(sub.dat, tax.id.type = "ncbi")
sigs <- BugSigDBStats::syncWithNCBI(sigs, onto)
sim.mat <- ontologySimilarity::get_sim_grid(ontology = onto, term_sets = sigs)
dist.mat <- stats::as.dist(1 - sim.mat)
res <- stats::cmdscale(dist.mat)
```

Now let's focus on fecal samples and signatures with at least 5 taxa.

```{r}
dat.feces <- subset(dat, `Body site` == "feces")
ind <- lengths(dat.feces[["NCBI Taxonomy IDs"]]) > 4
dat.feces <- dat.feces[ind,]
```

Pull out signatures based on NCBI Taxonomy IDs:

```{r}
sigs <- bugsigdbr::getSignatures(dat.feces, tax.id.type = "ncbi")
sigs <- BugSigDBStats::syncWithNCBI(sigs, onto) 
```

Compute semantic similarity:

```{r}
sim.mat <- ontologySimilarity::get_sim_grid(ontology = onto, term_sets = sigs)
sim.mat[1:5,1:5]
```

# Hierarchical clustering on the semantic similarity matrix

```{r}
hc <- stats::hclust(stats::as.dist(1 - sim.mat), method = "ward.D")
clus <- stats::cutree(hc, 5)
head(clus)
d <- data.frame(label = names(clus), count = lengths(sigs[names(clus)]))
head(d)
```

# Jaccard matrix on genus level

```{r}
sigs.genus <- bugsigdbr::getSignatures(dat.feces,
                                       #tax.id.type = "metaphlan",
                                       tax.id.type = "taxname", 
                                       tax.level = "genus",
                                       exact.tax.level = FALSE)
jmat <- BugSigDBStats::calcJaccardSimilarity(sigs.genus)
```

Correlation between semantic similarity and Jaccard similarity:

```{r}
.fcor <- function(i) cor(jmat[i,], sim.mat[rownames(jmat)[i], colnames(jmat)])
cors <- vapply(seq_len(nrow(jmat)), .fcor, numeric(1))
summary(cors)
```

# Heatmap comparison

Color ramp from 0.01 quantile to 0.99 quantile

```{r, fig.width = 10, fig.height = 10}
quantile(as.vector(sim.mat), 0.01)
quantile(as.vector(sim.mat), 0.99)
col2 <- circlize::colorRamp2(c(0,0.8012), c("#EEEEEE", "red"))

h2 <- ComplexHeatmap::Heatmap(sim.mat[rownames(jmat),rownames(jmat)], 
                              name = "semsim",
                              col = col2,
                              #column_km = 5,
                              #row_km = 6,
                              row_title = "signatures",
                              column_title = "signatures",
                              show_row_names=FALSE,
                              show_column_names=FALSE)

col <- circlize::colorRamp2(c(0,1/3), c("#EEEEEE", "red"))
h1 <- Heatmap(jmat[,column_order(h2)], 
              col = col,
              cluster_columns = FALSE,
              show_row_names = FALSE,
              show_column_names = FALSE,
              name = "jaccard")

h2 + h1
```

Zoom into a cluster:

```{r}
h2 <- draw(h2)
ro <- ComplexHeatmap::row_order(h2)
rd <- ComplexHeatmap::row_dend(h2)
str(rd, max.level = 2)
```

Cut dendrogram at specified height ...
```{r}
hd <- cut(rd, 2.90)
par(mar = c(0,0,4,20))
par(cex = 0.5)
plot(hd$lower[[1]], horiz = TRUE)
```

... or specified number of clusters:

```{r}
##  for hierarchical clustering
clus <- stats::cutree(as.hclust(rd), 5) 

## for k-means
#clus <- rep(seq_along(ro), lengths(ro))  
#names(clus) <- rownames(sim.mat)[unlist(ro)]
```

# Test each cluster for over-representation of a condition 

```{r}
cond <- .getCondition(names(clus))
tab <- table(cond) # background
spl <- split(cond, clus) # numbers per cluster
spl <- lapply(spl, table) # numbers per cluster
smat <- vapply(spl, function(x) x[names(tab)], integer(length(tab)))
smat[is.na(smat)] <- 0
rownames(smat) <- names(tab)
rs <- rowSums(smat)
cs <- colSums(smat)
total <- sum(tab)
```

Function for constructing 2x2 contingency table for one condition and one cluster at a time:

```{r}
getContingency <- function(co, cl, m, rs, cs, total) 
{
    both <- m[co,cl]
    one <- rs[co] - both
    two <- cs[cl] - both
    all <- total - one - two - both
    unname(c(both, two, one, all))
}

getContingency("COVID-19", 2, smat, rs, cs, total)
```

Build all condition/cluster combinations for testing:

```{r}
combs <- expand.grid(rownames(smat), seq_len(ncol(smat)))
conts <- apply(combs, 1, function(x) getContingency(x[1], x[2], smat, rs, cs, total)) 
conts <- t(conts)
rownames(conts) <- apply(combs, 1, function(x) paste(x, collapse = "."))
ind <- conts[,1] > 4
conts <- conts[ind,]
combs <- combs[ind,]
```

Test for over-representation using Fisher's exact test:

```{r}
fisherp <- function(x) fisher.test(matrix(x, nrow = 2), 
                       alternative = "greater")$p.value 
ps <- apply(conts, 1, fisherp)
adjp <- p.adjust(ps, method = "BH")
res <- data.frame(condition = combs[,1], 
                  cluster = combs[,2],
                  PVAL = ps,
                  ADJ.PVAL = adjp)
res <- res[order(ps),]
head(res, n = 10)
```

# Test each condition for semantic similarity

```{r}
dir <- .getDirection(rownames(sim.mat))
cond <- .getCondition(rownames(sim.mat))
ays <- vapply(rownames(sim.mat), .getAuthorYear, character(1), dat = dat) 
ays <- strsplit(unname(ays), "_")
ays <- vapply(ays, `[`, character(1), x = 1)
```

Select conditions studied in at least two studies:

```{r}
spl <- split(ays, cond)
spl <- lapply(spl, unique)
lens <- lengths(spl)
names(lens) <- names(spl)
lens <- sort(lens, decreasing = TRUE)
conds.to.test <- names(lens)[lens > 1] # 36 conditions
```

Function for testing signatures of a condition of interest:

```{r}
.testCondition <- function(condition, sim.mat, direction = "UP")
{
    ind <- which(cond == condition & dir == direction)
    if(length(ind) < 2) res <- c(sim = NA, p = NA)
    else res <- c(sim = ontologySimilarity::get_sim(sim.mat, group = ind),
                  p = ontologySimilarity::get_sim_p(sim.mat, group = ind))
    return(res)
}
```

```{r}
ps.up <- vapply(conds.to.test, .testCondition, numeric(2), sim.mat = sim.mat)
ps.up <- t(ps.up)
ps.up <- ps.up[!is.na(ps.up[,"p"]),]
ps.up <- ps.up[order(ps.up[,"p"]),]
df.up <- data.frame(CONDITION = rownames(ps.up), 
                    SEMSIM = ps.up[,"sim"],
                    PVAL = ps.up[,"p"],
                    NR.STUDIES = lens[rownames(ps.up)])
df.up
ps.down <- vapply(conds.to.test, .testCondition, numeric(2), sim.mat, direction = "DOWN")
ps.down <- t(ps.down)
ps.down <- ps.down[!is.na(ps.down[,"p"]),]
ps.down <- ps.down[order(ps.down[,"p"]),]
df.down <- data.frame(CONDITION = rownames(ps.down), 
                    SEMSIM = ps.down[,"sim"],
                    PVAL = ps.down[,"p"],
                    NR.STUDIES = lens[rownames(ps.down)])
df.down
```

Plot top consistent up and down conditions:

```{r}
isect <- intersect(rownames(df.up), rownames(df.down))
top.df <- cbind(df.up[isect,], df.down[isect,])
EnrichmentBrowser::configEBrowser("GS.COL", "CONDITION")
uranks <- EnrichmentBrowser:::.getRanks(df.up)
dranks <- EnrichmentBrowser:::.getRanks(df.down)
mranks <- (uranks[isect] + dranks[isect]) / 2 
ind <- order(mranks[rownames(top.df)])
top.df <- top.df[ind,]
colnames(top.df)[c(2:4,6:8)] <- paste(colnames(top.df)[c(2:4,6:8)], 
                                      rep(c("UP", "DOWN"), each = 3), 
                                      sep = ".") 
```

```{r}
df <- data.frame(CONDITION = rep(rownames(top.df), 2),
                 PVAL = -log10(c(top.df$PVAL.UP, top.df$PVAL.DOWN)), 
                 NR.STUDIES = c(top.df$NR.STUDIES.UP, top.df$NR.STUDIES.DOWN),   
                 DIRECTION = rep(c("Increased", "Decreased"), each = nrow(top.df)))
df <- df[c(1:20,30:49),]
df$CONDITION <- factor(df$CONDITION, levels = rev(df$CONDITION[1:20]))
df$DIRECTION <- factor(df$DIRECTION, levels = c("Increased", "Decreased"))
```

```{r, fig.height = 8}
dotp <- ggplot(df, aes_string(x = "PVAL", y = "CONDITION",
                              size = "NR.STUDIES", color = "PVAL")) + 
        geom_point() +
        scale_color_continuous(low = "grey", high = "red", name = "-log10(P)", 
                               guide = guide_colorbar(reverse = TRUE)) +
        scale_y_discrete(labels = enrichplot:::default_labeller(30)) +
        ggplot2::xlab("-log10(P)") +
        ggplot2::ylab(NULL) +
        ggplot2::ggtitle("") +
        enrichplot:::theme_dose() + 
        scale_size(range = c(3, 8)) +
        facet_wrap(vars(DIRECTION), ncol = 2)
dotp
```

# Test replicability across body sites

Get the data:

```{r}
df <- bugsigdbr::importBugSigDB(version = "10.5281/zenodo.5904281")
onto <- BugSigDBStats::getNcbiTaxonomyObo()
```

```{r}
ind <- !is.na(dat[["Body site"]]) & dat[["Body site"]] == "posterior fornix of vagina"
dat[ind, "Body site"] <- "vagina"
ind <- !is.na(dat$Condition) & dat$Condition == "eczema"
dat[ind, "Condition"] <- "atopic eczema"
res <- BugSigDBStats::testReplicability(dat,
                                        onto = onto,
                                        multi.contrast = "merge")
```

```{r}
res <- res[order(res$PVAL),]
top25 <- res[1:25,]
top25s <- paste(top25$CONDITION, top25[["BODY.SITE"]], sep = "-")
top25s <- unique(top25s) 
top25s <- c(paste(top25s, "UP", sep = "-"), paste(top25s, "DOWN", sep = "-"))
top25r <- paste(res$CONDITION, res[["BODY.SITE"]], res$DIRECTION, sep = "-")
top.df <- res[top25r %in% top25s,]
```

```{r}
top.df$CONDITION <- gsub("glandular intraepithelial", "", top.df$CONDITION)
top.df$CONDITION <- gsub("human papilloma virus", "HPV", top.df$CONDITION)
top.df$CONDITION <- gsub(" mellitus", "", top.df$CONDITION)
top.df[["BODY.SITE"]] <- gsub(" of body", "", top.df[["BODY.SITE"]])
top.df[["BODY.SITE"]] <- gsub("ine cervix", "us", top.df[["BODY.SITE"]])
top.df$PVAL <- -log10(top.df$PVAL)
```

```{r}
top.df$CONDITION <- paste0(top.df$CONDITION, " (", top.df[["BODY.SITE"]], ")") 
top.df$CONDITION <- factor(top.df$CONDITION, levels = rev(unique(top.df$CONDITION[1:25])))
top.df$DIRECTION <- factor(top.df$DIRECTION, levels = c("DOWN", "UP"))
top.df$DIRECTION <- ifelse(top.df$DIRECTION == "DOWN", "Decreased", "Increased")
top.df$DIRECTION <- factor(top.df$DIRECTION, levels = c("Decreased", "Increased"))
```

```{r, fig.height = 8}
dotp <- ggplot(top.df, aes_string(x = "PVAL", y = "CONDITION",
                              size = "NR.STUDIES", color = "PVAL")) + 
        geom_point() +
        scale_color_continuous(low = "grey", high = "red", name = "-log10(P)", 
                               guide = guide_colorbar(reverse = TRUE)) +
        scale_y_discrete(labels = enrichplot:::default_labeller(30)) +
        ggplot2::xlab("-log10(P)") +
        ggplot2::ylab(NULL) +
        ggplot2::ggtitle("") +
        enrichplot:::theme_dose() + 
        scale_size(range = c(3, 8), breaks = seq(5,30, by = 5)) +
        facet_wrap(vars(DIRECTION), ncol = 2)
dotp
```

Take a closer look at antibiotics and HIV infection:

Parameterization:

```{r}
multi.contrast <- "merge"
min.taxa <- 5
```

Sanity checks:

```{r}
stopifnot("Condition" %in% colnames(df))
stopifnot("Body site" %in% colnames(df))
stopifnot("NCBI Taxonomy IDs" %in% colnames(df))
if(!is.null(onto)) stopifnot(is(onto, "ontology_index"))
```

Clean pmid, condition, and body site:

```{r}
rel.cols <- c("PMID", "Condition", "Body site") 
for(column in rel.cols)
{
    ind <- !is.na(df[[column]]) & !grepl(",", df[[column]])
    df <- df[ind,]
}
```

```{r}
multi.bs <- length(unique(df[["Body site"]])) > 1
if(multi.bs) df$Condition <- paste(df$Condition, df[["Body site"]], sep = "*;*")
```

Treat multi contrast studies:

```{r}
df.up <- BugSigDBStats:::.restrictByDirection(df, direction = "UP")
df.down <- BugSigDBStats:::.restrictByDirection(df, direction = "DOWN")
df.up <- BugSigDBStats:::.treatMultiContrastStudies(df.up, multi.contrast)
df.down <- BugSigDBStats:::.treatMultiContrastStudies(df.down, multi.contrast)
df <- rbind(df.up, df.down)
```

```{r}
ind <- lengths(df[["NCBI Taxonomy IDs"]]) >= min.taxa
df <- df[ind,]
```

Calculate semantic similarity:

```{r}
sigs <- bugsigdbr::getSignatures(df, tax.id.type = "ncbi")
sigs <- BugSigDBStats::syncWithNCBI(sigs, onto)
sim.mat <- ontologySimilarity::get_sim_grid(ontology = onto, term_sets = sigs)
```

Get condition and direction of abundance change for each signature:

```{r}
cond <- .getCondition(names(sigs))
dir <- .getDirection(names(sigs))
```

Get authors and publication year for each signature:

```{r}
ays <- vapply(rownames(sim.mat), .getAuthorYear, character(1), dat = df)
ays <- strsplit(unname(ays), "_")
ays <- vapply(ays, `[`, character(1), x = 1)
head(ays)
```

Select signatures with decreased abundance upon antibiotics treatment:

```{r}
ind <- cond == "antimicrobial-agent*;*feces" & dir == "DOWN"
table(ind)
```

Get genus level signatures with taxname only for visualization purposes:

```{r}
sigs.genus <- bugsigdbr::getSignatures(df,
                                       tax.id.type = "taxname",
                                       tax.level = "genus",
                                       exact.tax.level = FALSE,
                                       min.size = 0)
```

Get also full metaphlan strings for analysis purposes:

```{r}
sigs.mp <- bugsigdbr::getSignatures(df, tax.id.type = "metaphlan", min.size = 0)
```

Get signature by genus incidence matrix: 

```{r}
psigs.mp <- sigs.mp[ind]
cmat <- safe::getCmatrix(psigs.mp, as.matrix = TRUE, min.size = 0, prune = FALSE)
```

Create signature labels for plotting:

```{r}
n <- c("infant:ampicillin+gentamicin",
       "elderly:antimicrobial-agent",
       "adult:GSK1322322",
       "adult:ciprofloxacin/clindamycin",
       "adult:moxifloxacin",
       "infant:prenatal-antibiotics",
       "infant:amoxicillin/penicillin",
       "infant:prenatal-antibiotics")
```

```{r}
psigs <- sigs.genus[ind]
smat <- sim.mat[ind,ind]
s <- strsplit(names(psigs), "_")
s <- vapply(s, `[`, character(1), x = 1)
s
```

```{r}
aysn <- ays[ind]
aysn <- sub("de", "Gunzburg", aysn)
aysn <- paste(aysn, n, sep = "-")
```

```{r}
names(psigs) <- rownames(smat) <- colnames(smat) <- aysn
hc <- stats::hclust(stats::as.dist(1 - smat), method = "ward.D")
freq <- table(unlist(psigs)) / length(psigs)
ddf <- data.frame(frequency = as.vector(freq))
rownames(ddf) <- names(freq)
```

Plot genus-level matrix:

```{r, fig.width = 10, fig.height = 10}
heatp1 <- EnrichmentPlotter::heatSimPlot(psigs, smat, ddf, "frequency") %>% 
          insert_left(ggtree(hc), width = 0.2)
heatp1
```

And network plot:

```{r}
np1 <- EnrichmentPlotter::networkSimPlot(psigs, smat, min.sim = 0.7)
np1
```

```{r}
snames <- c("Rashid2015", "Arat2015", "Gunzburg2018", "Zhang2019",
            "Coker2020", "Zou2018", "O'Sullivan2013", "Fouhy2012",
            "Laubitz2021")
```

Plot full taxon matrix:

```{r, fig.width = 10, fig.height = 10}
col.fun <- circlize::colorRamp2(c(0, 1), c("white", "black"))
spl <- strsplit(aysn, "-")
spl <- vapply(spl, `[`, character(1), x = 1)
colnames(cmat) <- spl
rownames(cmat) <- sub("k__Bacteria\\|", "", rownames(cmat))
ComplexHeatmap::Heatmap(cmat[,snames],
                        col = col.fun,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                        row_names_gp = gpar(fontsize = 7),
                        column_names_gp = gpar(fontsize = 8),
                        row_names_max_width = unit(15, "cm"),
                        show_heatmap_legend = FALSE)
```

Select signatures with decreased abundance upon HIV-infection:

```{r}
ind <- cond == "HIV-infection*;*feces" & dir == "DOWN"
table(ind)
```
Get signature by genus incidence matrix: 

```{r}
psigs.mp <- sigs.mp[ind]
cmat <- safe::getCmatrix(psigs.mp, as.matrix = TRUE, min.size = 0, prune = FALSE)
```

Create signature labels for plotting:

```{r}
n <- c("adult:HIV-infection",
       "adult:HIV-infection",
       "adult:HIV-infection",
       "children:perinatal-HIV-infection",
       "children:perinatal-HIV-infection")
```

```{r}
psigs <- sigs.genus[ind]
smat <- sim.mat[ind,ind]
s <- strsplit(names(psigs), "_")
s <- vapply(s, `[`, character(1), x = 1)
s
```

```{r}
aysn <- ays[ind]
aysn <- paste(aysn, n, sep = "-")
aysn
```

```{r}
names(psigs) <- rownames(smat) <- colnames(smat) <- aysn
hc <- stats::hclust(stats::as.dist(1 - smat), method = "ward.D")
freq <- table(unlist(psigs)) / length(psigs)
ddf <- data.frame(frequency = as.vector(freq))
rownames(ddf) <- names(freq)
```

Plot genus-level matrix:

```{r, fig.width = 10, fig.height = 10}
heatp2 <- EnrichmentPlotter::heatSimPlot(psigs, smat, ddf, "frequency") %>% 
          insert_left(ggtree(hc), width = 0.2)
heatp2
```

And network plot:

```{r}
np2 <- EnrichmentPlotter::networkSimPlot(psigs, smat, min.sim = 0.5)
np2
```

```{r}
snames <- c("McHardy2013", "Ling2016", "Zhou2018", "Kaur2018", "Moya2020")
```

Plot full taxon matrix:

```{r, fig.width = 10, fig.height = 10}
col.fun <- circlize::colorRamp2(c(0, 1), c("white", "black"))
spl <- strsplit(aysn, "-")
spl <- vapply(spl, `[`, character(1), x = 1)
colnames(cmat) <- spl
ComplexHeatmap::Heatmap(cmat[,snames],
                        col = col.fun,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                        row_names_gp = gpar(fontsize = 7),
                        column_names_gp = gpar(fontsize = 8),
                        row_names_max_width = unit(14, "cm"),
                        show_heatmap_legend = FALSE)
```
