---
title: Table 1. BugSigDB metadata annotations 
vignette: >
  % \VignetteIndexEntry{Figure 4}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output:
  html_document:
    mathjax: null   
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
library(bugsigdbr)
library(BugSigDBStats)
```

# Get bulk export (version 1.0.2)

```{r}
dat <- bugsigdbr::importBugSigDB(version = "10.5281/zenodo.5904281", cache = FALSE)
dim(dat)
```

# Study stats

## Study design

```{r}
spl <- split(dat[["Study"]], dat[["Study design"]])
sds <- lapply(spl, unique)
sort(lengths(sds), decreasing = FALSE)
```

# Experiment stats

Columns of the full dataset that describe experiments:

```{r}
# Experiment ID
exp.cols <- c("Study", "Experiment")

# Subjects
sub.cols <- c("Host species",
              "Location of subjects",
              "Body site",
              "Condition",
              "Antibiotics exclusion",
              "Group 0 sample size",
              "Group 1 sample size")

# Lab analysis              
lab.cols <-  c("Sequencing type",
              "16S variable region",
              "Sequencing platform")

# Statistical analysis
stat.cols <-  c("Statistical test",
              "MHT correction",
              "Significance threshold")

# Alpha diversity
div.cols <- c("Pielou",
              "Shannon",
              "Chao1",
              "Simpson",
              "Inverse Simpson",
              "Richness")
```

Restrict dataset to experiment information:

```{r}
exps <- dat[,c(exp.cols, sub.cols, lab.cols, stat.cols, div.cols)]
exps <- unique(exps)
```

## Subjects

Number of experiments for the top 10 categories for each subjects column:

```{r}
sub.tab <- lapply(sub.cols[1:5], tabCol, df = exps, n = 10)
names(sub.tab) <- sub.cols[1:5]
sub.tab
```

Proportions instead:

```{r}
sub.tab <- lapply(sub.cols[1:5], tabCol, df = exps, n = 10, perc = TRUE)
names(sub.tab) <- sub.cols[1:5]
sub.tab
```

Sample size:

```{r}
ssize <- apply(exps[,sub.cols[6:7]], 2, summary)
ssize
```

## Lab analysis

Number of experiments for the top 10 categories for each lab analysis column:

```{r}
lab.tab <- lapply(lab.cols, tabCol, df = exps, n = 10)
names(lab.tab) <- lab.cols
lab.tab
```

Proportions instead:

```{r}
lab.tab <- lapply(lab.cols, tabCol, df = exps, n = 10, perc = TRUE)
names(lab.tab) <- lab.cols
lab.tab
```

## Statistical analysis

Number of experiments for the top 10 categories for each statistical analysis column:

```{r}
stat.tab <- lapply(stat.cols, tabCol, df = exps, n = 10)
names(stat.tab) <- stat.cols
stat.tab
```

Proportions instead:

```{r}
stat.tab <- lapply(stat.cols, tabCol, df = exps, n = 10, perc = TRUE)
names(stat.tab) <- stat.cols
stat.tab
```

## Alpha diversity

Overall distribution:

```{r}
apply(exps[,div.cols], 2, table)
```

Correspondence of Shannon diversity and Richness:

```{r}
table(exps$Shannon, exps$Richness)
```

Conditions with consistently increased or decreased alpha diversity:

```{r}
BugSigDBStats::tabDiv(exps, "Shannon", "Condition")
BugSigDBStats::tabDiv(exps, "Shannon", "Condition", perc = TRUE)
BugSigDBStats::tabDiv(exps, "Richness", "Condition")
BugSigDBStats::tabDiv(exps, "Richness", "Condition", perc = TRUE)
BugSigDBStats::tabDiv(exps, "any", "Condition")
BugSigDBStats::tabDiv(exps, "any", "Condition", perc = TRUE)
```

Body sites with consistently increased or decreased alpha diversity:

```{r}
BugSigDBStats::tabDiv(exps, "Shannon", "Body site")
BugSigDBStats::tabDiv(exps, "Shannon", "Body site", perc = TRUE)
BugSigDBStats::tabDiv(exps, "Richness", "Body site")
BugSigDBStats::tabDiv(exps, "Richness", "Body site", perc = TRUE)
BugSigDBStats::tabDiv(exps, "any", "Body site")
BugSigDBStats::tabDiv(exps, "any", "Body site", perc = TRUE)
```

How many experiments do not report alpha diversity:
 
```{r}
ind <- apply(exps[,div.cols], 1, function(x) all(is.na(x)))
table(ind)
```

How many experiments report at least one increased / decreased measure of alpha 
diversity, and how many do not report any change in alpha diversity:

```{r}
rexps <- exps[!ind,div.cols]
x <- apply(rexps, 1, table)
f <- function(s)
{
    s2 <- s
    if(length(s) > 1)
    {
        s2 <- s[names(s) != "unchanged"]
        s2 <- s2[which.max(s2)]
    }
    return(names(s2))
} 
y <- vapply(x, f, character(1))
table(y)
table(y) / nrow(exps) 
```

